Cohomology Theories, Triangulated Categories, and Applications\vspace{-1ex}

### 2019-06-05 Homological Algebra Notes/Hom Alg.tex
Almost the end!

\section{June 3, 2019}
Today we're talking about spectral sequences! The primary references are Vakil's ``The Rising Sea'' (specifically section 1.7)
as well as Weiel and May's ``A primer on spectral sequences.''

@@

>>>>

\section{June 3, 2019}
Today we're talking about spectral sequences! Sean Griffin is presenting. The primary references are Vakil's ``The Rising Sea'' (specifically section 1.7)
as well as Weibel and May's ``A primer on spectral sequences.''

@@

\section{June 5, 2019}
Today we are going to discuss Galois cohomology. The Thomases Carr and Browning are presenting.

\subsection{Getting Started}
The set up here is that we have an extension $L/K$ of fields with $G=\operatorname{Gal}(L/K)$. It ends up that really Galois cohomology is 
really just group cohomology, transported to this context. There are two modules one may consider: $L^+$ and $L^\times$. Unfortunately,
\begin{thm}
	$H^i(G,L^+)=0$ for all $i>0$.
\end{thm}
So instead we ocus on using $L^\times.$ Here we still get a nice result:
\begin{thm}[Hilbert Theorem 90]
	$H^1(G,L^\times)=1$ 
\end{thm}
\begin{rmk}
	Notice that in general the higher $H^i(G,L^\times)$ may or may not be trivial.
\end{rmk}
\subsection{Some Definitions}
\begin{defn}
	An abelian extension $L/K$ with Galois group $G$ is said to be of \textbf{exponent $n$} if $\sigma^n=1$ for all $\sigma\in G$.
\end{defn}
\begin{defn}
	A \textbf{Kummer extension} $L/K$ is an abelian extension of exponent $n$ where $k$ contains all $n^{th}$ roots of unity.
\end{defn}
\begin{ex}
	When $n=2$, $L=K(\sqrt{a})$ where $a$ is a nonsquare in $K$ is a Kummer extension.

	When $n=3$, we can let $L$ be the splitting field of $x^3-\alpha$ where $\alpha$ is a non-cube. Then we set $K$ to be the subfield containing the third roots of unity.
\end{ex}

\subsection{Good Stuff}
Let $L/K$ be Kummer. Consider 
\[1\to \mu_n\to L^\times\xrightarrow{n^{th}\text{power}}(L^\times)^n)\to 1\]
and using Galois cohomology, we get the long exact sequence (leveraging that $H^1(G,L^\times)=1$
\[1\to \mu_n\to K^\times\to K^\times\cap(L^\times)^n\twoheadrightarrow H^1(G,\mu_n)\to 1\]
and so since $H^1(G,\mu_n)$ ($\mu_n$ is a trivial module and $G$ is abelian although the latter is not necessary), 
\[K^\times\cap(L^\times)^n/(K^\times)^n\cong \Hom(G,\mu_n).\]

But then we can use a form of Pontryagin duality,
\[\Hom(G,\mu_n)\cong\Hom(G,\bbZ/\bbZ)\]
There is a

And we get an overall correspondence between subgroups of $K^\times\cap (L^\times)^n$ containing $(K^\times)^n$ and 
Kummer extensions $L/K$ of exponent $n$. The upshot here is that if we pass to some maximal Kummer extension of $K$, (call it $N$), 
then $K^\times\cap (N^\times)^n=K^\times$, so subgroups of $K^\times$ containing $(K^\times)^n)$ are in bijection with 
Kummer extensions of $K$ of exponent $n$.

More explicitly the correspondence takes a field $L$ to a group $(L^\times)^n\cap K^\times$ and a group $\Delta$ to $K(\sqrt{\Delta})=K(\sqrt[n]{\alpha},\alpha\in\Delta)$.

\subsection{What this means}
As an application, we get a correspondence between the subgroups of $\bbQ^\times/(\bbQ^\times)^2$ and Kummer extensions of exponent 2.

\subsection{Global Stuff}
Let $K$ be a number field (a finite extension over $\bbQ$). Then we get a SES 
\[1\to K^\times\to I_K\to C_K\to 1\]
where $I_K$ is the so-called \textbf{idele group} and $C_K$ is the \textbf{idele class group}.

To define these we need some fancy stuff so I am going to ignore it. Something about restricted products. :)

Now if $L/K$ is Galois, we get $K^\times\subseteq L^\times$. This naturally gives us $I_K\subseteq I_L$ and some kind of inclusion $C_K\to C_L$.
If $G$ is the Galois group for $L/K$, then 
\[(L^\times)^G=K^\times\quad (I_L)^G=I_K\quad (C_L)^G=C_K.\]

Introduce the following notation:
\[H^n(L/K)=H^n(\operatorname{Gal}(L/K),C_L)\]
Then we get the following generalization of Hilbert 90:
\begin{thm}
	$H^2(L/K)=1$
\end{thm}
\begin{prf}
	This is at least the interesting bit. Induct of $[L:K]$. Let $\operatorname{Gal}(L/F)$ be a Sylow $p$-subgroup 
	of the bigger Galois group. Then we get restriction and corestriction maps:
	\[H^1(L/K)\xrightarrow{res}H^1(L/F)\xrightarrow{cores}H^1(L/K)\]
	and in particular $res$ is injective on the $p$-subgroup. So our goal will be to show that $H^1(L/F)$ is trivial, 
	proving that the $p$ part of $H^1(L/K)$ is trivial, and letting it vary over primes.

	Now assume that $\operatorname{Gal}(L/K)$ to be a $p$-group and $\operatorname{Gal}(L/F)$ be normal of index $p$.
	Then you get an inflation restriction sequence 
	\[1\to H^1(F/K)\xrightarrow{inf}H^1(L/K)\xrightarrow{res}(L/F)\]
	and then you can do some number theory to prove what you want.
\end{prf}
\begin{rmk}
	Notice that the above sequence is, in fact, the first three terms of the five term exact sequence you get from the Lyndon-Hochschild-Serre
	spectral sequence in group cohomology.
\end{rmk}

\begin{thm}
	$H^2(L/K)\cong\left(\frac{1}{[L:K]}\bbZ\right)/\bbZ\subseteq \bbQ/\bbZ$ is cyclic of order $[L:K]$
\end{thm}

\subsection{A slight aside: Tate cohomology}
Let $G$ be a finite group and $A$ a $G$-module. Let $N=\sum_{g\in G} g\in\bbZ G$. Then we get a map 
$A_G\xrightarrow{N_G}A^G$. Then define

\[\hat H^n(G,A)=\left\{\begin{array}{lr} H^n(G,A),& n\ge 1\\
	\coker N_G, & n=0\\
	\ker N_G & n=-1\\
H_{-n-1}(G,A),& n\le -2\end{array}\right.\]

It just happens that this ends up working together very nicely!

\begin{prop}
	Given $0\to A\to B\to C\to 0$ an SES, we get our long exact sequence in Tate cohomology.
\end{prop}
\begin{prop}
	When $G$ is cyclic, $\hat H^n(G,A)\cong\hat H^{n+2}(G,A)$
\end{prop}
\begin{prop}
	We have a nice cup product
	\[\hat H^m(G,A)\otimes \hat H^n(G,B)\to \hat H^{n+m}(G,A\otimes B)\]
\end{prop}
\begin{prop}
	For all $m$ there exists a ``shift module'' $S_m$ such that 
	\[\hat H^n(G,A)\cong\hat H^{n-m}(G,A\otimes S_m).\]
\end{prop}
\begin{thm}[Tate]
	Let $G$ be finite, $A$ a $G$ module such that 
	\begin{itemize}
		\item $\forall H\le G,\hat H^1(H,A)=0$
		\item $\forall H\le G, \hat H^2(H,A)$ is cyclic of order $|H|$
	\end{itemize}
	Then if $A$ generates $\hat H^2(G,A)$ then we get an isomorphism
	\[a\cup -:\hat H^n(G,\bbZ)\cong \hat H^{n+2}(G,A)\]
\end{thm}
Then setting $n=-2$, we get that 
\[\hat H^{-2}(G,\bbZ)\cong\hat H^0(G,A)=\coker N_G=A^G/NA\]
but then 
\[A^G/NA\cing \hat H^0(G,A)\cong H_2(G,\bbZ)=G^{ab}\]
where the last thing is the abelianization of $G$.

$C_L^G/NC_L\cong\operatorname{Gal}(L/K)^{ab}$

### 2019-06-05 Triangulated Categories/Tri Cat.tex
Almost the end!

	(2)$\Rightarrow$(1) By the first lemma, we have $\Ext_A^i(k,A)=0$ when $i\ne d$ and is finite dimensional otherwise.
	Let $F(-)=\Ext_A^d(-,A)$ and $G(-)=\Ext^d_{A^{op}}(-,A)$. Now note that 
	\[F(M)=R\Hom_A(M,A)[-d]=R\Hom_A(M,A[d])\]
	and let $R=A[d]$. The claim is that $R$ is a dualizing complex. This implies that $F(M)=\bbD(M)$ and $G(N)=\bbD^\circ(N)$,
	and we can apply the lemma to show that $\Ext_A^d(k,A)=F(k)=k$ and $\Ext_{A^{op}}^d(k,A)=k$, so we're done.
\end{prf}

>>>>

	(2)$\Rightarrow$(1) By the first lemma, we have $\Ext_A^i(k,A)=0$ when $i\ne d$ and is finite dimensional otherwise.
	Let $F(-)=\Ext_A^d(-,A)$ and $G(-)=\Ext^d_{A^{op}}(-,A)$. Now note that 
	\[F(M)=R\Hom_A(M,A)[-d]=R\Hom_A(M,A[d])\]
	and let $R=A[d]$. The claim is that $R$ is a dualizing complex. This implies that $F(M)=\bbD(M)$ and $G(N)=\bbD^\circ(N)$,
	and we can apply the lemma to show that $\Ext_A^d(k,A)=F(k)=k$ and $\Ext_{A^{op}}^d(k,A)=k$, so we're done.
\end{prf}	
\begin{rmk}
	In all honesty I lost track of the proof near the end, but I am pretty confused about whether that last part actually proved the equivalence of 4 or not.
	In class the next day, James mentioned that (4)$\Rightarrow$(2) is a result of Yekutieli in his thesis.
\end{rmk}

\section{June 3, 2019}
Today we are talking about Poincar\'e duality and something about Nakayama.
\subsection{Examples of AS regular algebras}
Examples abound, but this result gives us several:
\begin{thm}[Brown]
	Let $A$ be a connected graded algebra over a field of characteristic zero with finite GK dimension (think noncommutative Krull dimension). If $A$ admits a Hopf 
	algebra structure, then $A$ must be AS regular.

	Furthermore, $A$ satisfies 
\[\Ext_{A^e}(A,A^e)=\left\{\begin{array}{rl}
	0, & \text{GK dim}\ne i\\
	A,&\text{otherwise}\end{array}\right.\]
\end{thm}

\subsection{Some Major Theorems}
This one is courtesy of James and our guest Dan Rogalski!
\begin{thm}[RRZ]
	Let $A$ be connected graded. Then the following are equivalent:
	\begin{itemize}
		\item $A$ is AS regular.
		\item $A$ is skew Calabi-Yau.
	\end{itemize}
\end{thm}
Another from Michelle Van den Bergh:
\begin{thm}[Van den Bergh]\label{thm-VdB}
	Let $A$ be connected graded and AS regular (or equivalently skew Calabi-Yau). Then for any bimodule $M$,
	and for some automorphism $\mu$ of $A$,
	\[\HH^i(M)=\Ext^i_{A^e}(A,M)\cong \Tor_{d-i}^{A^{e}}(A,{^\mu}M)=\HH_{d-i}({^\mu}M)\]
\end{thm}
\begin{rmk}
	This gives us a form of Poincar\'e duality that is similar to the one that comes from topology.
\end{rmk}
\begin{defn}
	Let $M$ be an $A$ bimodule and $\mu$ an automorphism of $A$. Then $'M^\mu$ is a new $A$ bimodule where $A$ 
	has the same action on the left as $M$ but the right action is twisted by $\mu$.
\end{defn}
\begin{defn}
	An algebra $A$ is called skew Calabi-Yau (CY) if 
	\begin{itemize}
		\item $A\in\Dbperf(A^{e}\text{-}\mathbf{mod})$
		\item There is an integer $d$ and algebra automorphism $\mu$ such that 
	\[\Ext_{A^{e}}(A,A^e)=\left\{\begin{array}{rl}
	0, & i\ne d\\ 'A^\mu, & \text{otherwise}\end{array}\right.\]
	\end{itemize}
	Here $\mu$ is called a \textbf{Nakayama automorphism} of $A$.
\end{defn}
\begin{rmk}
	In 1988 Kansvitch introduced CW categories. Later in 2006 Gingzburg defined CYalg, the Calabi-Yau category. This 
	was done in a (mysteriously unpublished) paper that James highly recommends we read.
\end{rmk}
\begin{lem}\label{lem-useful}
	If $X\in\Dbperf(B)$, add $Y\in D(B)$, then 
	\[R\Hom_B(X,Y)\cong R\Hom_B(X,B)\otimes_B^LY.\]
\end{lem}
\begin{prf}
	The above equation holds when we set $X=B$. This is more-or-less self-evident.

	Then we can take $\calT$ to be the set of objects satisfying the above equation, show it is a triangulated subcategory, then show it's everything 
	because $\langle B\rangle=\Dbperf(B)$.
\end{prf}
\begin{prf}[of theorem~\ref{thm-VdB}]
	\begin{align*}
		\HH^i(M)&=H^i(R\Hom_{A^e}(A,M))\\
		&=H^i(R\Hom(A,A^e)\otimes_{A^e}^LM)\\
		&=H^i('A^\mu[-d]\otimes_{A^e}^LM)\\
		&=H^{i-d}('A^\mu\otimes_{A^e}^LM)\\
		&=H^{i-d}(A\otimes_{A^e}^L(A\otimes 'A^\mu)\otimes_{A^e}^LM)\\
		&=H^{i=d}(A\otimes_{A^e}^L({^\mu}M))\\
		&=\HH_{d-i}(^{\mu}M).
	\end{align*}
\end{prf}
\subsection{Another neat result}
\begin{thm}[S. P. Smith]
	Let $A$ be a connected graded algebra. Then $A$ is AS regular if and only if $\Ext_A^\ast$ is Frobenius.
\end{thm}

Next lecture will be our last class and we will discuss the proof to James' theorem mentioned above.

\section{June 5, 2019}
Today we will finish up establishing that AS regular algebras are precisely the skey CY ones.

As a special calse, say that $L$ is an $N_{\ge 0}$-graded Lie algebra what is finite dimensional over $k$. Then $\calU(L)$ is connected graded and 
furthermore is AS regular and CY.

\begin{defn}
	An algebra $A$ is called \textbf{skew Calabi-Yau} if 
	\begin{itemize}
		\item $A\in\Dbperf(A^{e}\text{-}\mathbf{mod})$
		\item $\HH^i(A^e)=\Ext_{A^e}^i(A,A^e)=\delta_{i=d}{'}A^\mu$.
	\end{itemize}
\end{defn}

To prove [RRZ], we need several lemmas. (10 in total although we only need four to prove (skew CY)$\Rightarrow$(AS)).
\begin{lem}
	\[\gldim A\le \projdim_{A^e}A\]
\end{lem}
\begin{lem}
	\begin{itemize}
		\item $\gldim A\otimes B\le \projdim_{A^e\otimes B^e}(A\otimes B)$
		\item $\projdim_{A^e}(A)\le \gldim A^e\le \projdim_{(A^e)^e)}A^e=2\projdim_{A^e}A.$
	\end{itemize}
\end{lem}
Then we also have lemma~\ref{lem-useful} from yesterday, as well as 
\begin{lem}
	Let $\sigma$ be an automorphism of $A$. Then 
	\begin{itemize}
		\item $('A^\sigma\otimes k)\otimes_{A^e}^LA\cong k$
		\item $A\otimes_{A^e}^L('A^\sigma\otimes k)\cong k$.
	\end{itemize}
\end{lem}
\begin{prf}
	Let $P$ be a projective resolution of ${_{A^e}}A$. Then 
	\[('A^\sigma\otimes k)\otimes_{A^e}^LA=('A^\sigma\otimes k)\otimes_{A^e}P='A^\sigma\otimes_AP\otimes_Ak={^{\sigma^{-1}}}P\otimes_A k={^{\sigma^{-1}}}A\otimes_Ak={^{\sigma^{-1}}}k\]
\end{prf}

\subsection{Koszul Duality}
\begin{defn}
	Let $A$ be a connected graded algebra that is generated in degree 1. Then $A$ is called \textbf{Koszul} if 
	(there are several equivalent definitions but this is the easiest) 
	\[E:=\oplus_{i\ge 0}\Ext_A^i(k,k)\] 
	is generated by $\Ext_A^1(k,k)$.
\end{defn}
\begin{thm}[Koszul Duality: Beilinson-Ginzburg-Soergel]
	let $A$ be a Koszul AS-regular algebra. Then we have an equivalence of categories
	\[\Db_{fg}(\Amod)\simeq \Db_{fg}(E\text{-}\mathbf{mod}).\]
\end{thm}
\begin{cor}
	$\HH^\ast(A)\cong\HH^\ast(E)$.
\end{cor}
\begin{rmk}
	Notice that we \textbf{do not} get isomorphism on the individual graded pieces above: in general, $\HH^i(A)\not\cong\HH^i(E)$.
	In fact, there are only finitely many $\HH^i(A)$ nonzero, while this is not true for $\HH^i(E)$.
\end{rmk}

\subsection{The right context for noncommutative algebraic geometry}
\begin{defn}
	$\mathbf{proj} A$ is the category of graded modules over $A$ modulo the finitely generated ones. This is an abelian category.
\end{defn}
\begin{thm}[Beilinson-Gelfand-Gelfand]
	$\Db(\mathbf{proj} A)\cong \Db_{fg}(E\text{-}\textbf{Grmod})/\Dbperf(E\text{-}\mathbf{mod}).$
\end{thm}

\begin{thm}[Beilinson]
	$\Db(\mathbf{proj} A)\cong\Db_{fd}(B\text{-}\mathbf{mod})$ where $B\subseteq M_{d\times d}(A)$
	where $d=\gldim A$, where the matrices look like elements of $k$ down the diagonal and when $j>i$, the element in 
	the $(i,j)^{th}$ position comes from $A_{j-i}.$
\end{thm}

\begin{thm}[Orlov]
	Let $\bbV=\Db_{fd}(\Vectk)$. Then 
	\[\Db(\mathbf{proj}A)\cong (\bbV\rtimes \bbV)\rtimes\bbV\cdots\rtimes \bbV\]
	where the above is $d=\gldim A$ copies of $A$. We haven't defined the semidirect product of a category, but it can be done. This is called a semi-orthogonal decomposition of $\Db(\mathbf{proj}A)$.
\end{thm}
